# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data  
The first step is to read the data. The first several rows of the resulting data frame are printed, so the reader can see the form of the data.
```{r}  
pa1data <- read.csv("activity.csv")  
head(pa1data, 5)
```  

## What is mean total number of steps taken per day?
To summarise the data and get an idea of the result to be expected, we make a histogram of the total number of steps taken in a day (excluding missing values).
```{r, fig.width=8,fig.height=8}
totalsteps <- aggregate(pa1data$steps,by=list(pa1data$date),sum, na.rm=TRUE)
names(totalsteps) <- c("date", "totalSteps")
hist(totalsteps$totalSteps, xlab="Total steps", main="",breaks=seq(0,floor(max(totalsteps$totalSteps)/1000+1)*1000,1000))
```
Now, calculate the mean and median total number of steps taken per day using the aggregated data produced for the histogram.
```{r}
mean(totalsteps$totalSteps, na.rm=TRUE)
```

```{r}
median(totalsteps$totalSteps, na.rm=TRUE)
```

## What is the average daily activity pattern?
First, make a time series plot of the mean number of steps taken by interval (averaged over all days for that interval, with missing values removed).
```{r, fig.width=8,fig.height=8}
meanstepsbyinterval <- aggregate(pa1data$steps,by=list(pa1data$interval),mean,na.rm=TRUE)
names(meanstepsbyinterval) <- c("interval","meanStepsByInterval")
plot(meanstepsbyinterval$interval,meanstepsbyinterval$meanStepsByInterval,type="l",xlab="Interval",ylab="Mean steps taken",main="Daily Activity Pattern")
```
It looks like on average the most activity is in one of the intervals between 5:00AM and 10:00AM. Let's find out exactly which.
```{r}
meanstepsbyinterval[meanstepsbyinterval$meanStepsByInterval==max(meanstepsbyinterval$meanStepsByInterval),]
```
So, the most activity is in the interval from 8:35AM-8:40AM, when the person takes on average 206.2 steps.
## Imputing missing values
Let's figure out the total number of intervals for which the number of steps is missing (i.e. the number of rows with NAs).
```{r}
nrow(pa1data[!complete.cases(pa1data),])
```
That's a lot of missing data! Let's create a new dataset that replaces all of the NAs with the corresponding mean value for the interval in question, averaged over all days.

```{r}
completedata <- pa1data
completedata[is.na(completedata$steps),"steps"] <- meanstepsbyinterval[match(completedata[is.na(completedata$steps),"interval"],meanstepsbyinterval$interval),"meanStepsByInterval"]
```
Okay, now just like we did with the original data, we plot a histogram of the total number of steps taken in a day, and calculate the mean and median as well.
```{r, fig.width=8, fig.height=8}
completetotalsteps <- aggregate(completedata$steps,by=list(completedata$date),sum)
names(completetotalsteps) <- c("date", "totalSteps")
hist(completetotalsteps$totalSteps, xlab="Total steps", main="",breaks=seq(0,floor(max(completetotalsteps$totalSteps)/1000+1)*1000,1000))
m <- mean(completetotalsteps$totalSteps)
m
median(completetotalsteps$totalSteps)
```
So now, both the mean and median have increased from their previous values. As for the histogram, the frequency of the 0-1000 bin has noticeably decreased, and the frequency of the 10000-11000 bin has noticeably increased. The main reason for this is my choice to replace the NAs in the original data with the means for the same interval averaged over all days where a count was reported. A consequence of this is that for days where previously all of the step counts were NAs, the new step counts are equal to the sum of the interval means over all intervals, which is the same as the overall mean (`r format(m,digits=5)` when rounded to the nearest integer). The table below indicates the prevalence of the step count `r format(m,digits=5)` in the rounded data.
```{r}
sort(table(round(completetotalsteps$totalSteps)),decreasing=TRUE)
```
## Are there differences in activity patterns between weekdays and weekends?
